---
title: "Phylogenetic Tree Construction with FastTree2"
author: "Shawn Xinyu Wang"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document: 
    code_folding: show
    theme: spacelab
    highlight: pygments
    keep_md: no
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
      toc_depth: 3
  keep_md: true  
editor_options: 
  chunk_output_type: console
---


# Goals 

The goal of this script is to make a phylogenetic tree so we can use it for phylogenetic community analyses like phylogenetic Hill Numbers and the UniFrac beta-diversity measures. 

1. Load in pre-processed phyloseq object. 
2. Create ASV fasta file from the phyloseq object.
3. Align the 16S Sequences from our fasta file using the software [MAFFT](https://academic.oup.com/mbe/article/30/4/772/1073398).  
4. Create a maximum likelihood tree using the software [FastTree2](https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0009490). 
5. Write out the tree file for use in the next step of the analyses. 

## Inputs 
1. Our previously pre-processed phyloseq object that has chloroplasts, mitochondria, reverse complements, and negative controls removed, which is located in `data/03_PreProcessing/raw_preprocessed_physeq`

## Outputs 
1. The **aligned 16S rRNA gene sequences**, which is the input to the phylogenetic tree. We will save our alignment to `data/04_PhylogeneticTree/MAAFT_aligned_ASV.fasta`
2. The **calculated phylogenetic tree file**, which is the input to the phylogenetic tree. We will save our alignment to`data/04_PhylogeneticTree/ASVs_unrooted.tree`


# Setup 

## Timing of this script
Let's record how long this file took to run on the class server, which we will record at the end of the script. 
```{r rmd-start}
# What time did we start running this script? 
start_time <- Sys.time()
```

## Set Seed 
```{r set-seed}
set.seed(20020206)
```

## Load Packages 
```{r load-packages}
# Install phytools
#remotes::install_github("liamrevell/phytools")

# Install ggtree
#devtools::install_github("GuangchuangYu/ggtree")

# Now load them into the session 
pacman::p_load(phytools, ggtree, tidyverse, install = FALSE)
```

## 1. Load Data 

```{r load-physeq}
# Load physeq 
load("data/03_PreProcessing/raw_preprocessed_physeq.RData")
raw_preprocessed_physeq
```

# 2. Write ASV Fasta 

```{r asv-fasta}

# Biostrings provides DNAStringSet and FASTA export
library(Biostrings)

# Read the ASV abundance table (not transposed) from file
asv_table_raw <- read.delim("data/01_DADA2/ASV_table_raw.tsv",
                            sep = "\t", check.names = FALSE)

# Extract DNA sequences from the column names
asv_seqs <- colnames(asv_table_raw)

# Create ASV IDs corresponding to the sequences
asv_ids <- sprintf("ASV_%04d", seq_along(asv_seqs))

# Create DNAStringSet object with ASV IDs and sequences
asv_dna <- DNAStringSet(asv_seqs)
names(asv_dna) <- asv_ids

# Write the FASTA file to disk for MAFFT alignment
writeXStringSet(asv_dna,
                filepath = "data/04_PhylogeneticTree/preprocessed_ASVs.fasta")

```


# 3. Run Alignment with MAAFT
```{r run-mafft, engine = 'bash', engine.opts = '-l'}
# Write bash code to run MAFFT
# First provide the path to MAFFT
export PATH=/programs/mafft/bin:$PATH

# Where am I? 
echo "The working directory is $PWD"

# Set a seed for consistency and reproducibility 
RANDOM=20020206

# Now, actually run MAFFT
/programs/mafft/bin/mafft --auto data/04_PhylogeneticTree/preprocessed_ASVs.fasta > data/04_PhylogeneticTree/MAFFT_aligned_ASVs.fasta
```


# 4. Calculate the phylogenetic tree with FastTree2
```{r run-fasttree2, engine = 'bash', engine.opts = '-l'}
# Where am I? 
echo "The working directory is $PWD"

# Load fasttree 
# Provide export path to fasttree 
export PATH=/programs/FastTree-2.1.11:$PATH

# Run Fasttree to generate phylogenetic tree 
# parameters: 
    # -nt = indicates it's a nucleotide alignment
    # -gtr = generalized time reversible substitution model 
    # -fastest speed up the model, reduce memory usage (recommended for datasets that have >50,000)
    # -log = output a log file 
    # input alignment file 
    # specify the output tree file 
FastTree -nt -gtr -fastest -log data/04_PhylogeneticTree/FastTree.log data/04_PhylogeneticTree/MAFFT_aligned_ASVs.fasta > data/04_PhylogeneticTree/ASVs_unrooted.tree
```


# Final info for Reproducibility 

## Check Render Time
```{r stop-time}
# Take the time now that we are at the end of the script
end_time <- Sys.time()
end_time 

# Echo the elapsed time
elapsed_time <- round((end_time - start_time), 3)
elapsed_time
```


## Session Information
```{r session-info}
# Ensure reproducibility with package version information
devtools::session_info()
```




















