---
title: "Biodiversity across a salinity gradient"
author: "Mar Schmidt"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document: 
    code_folding: show
    theme: spacelab
    highlight: pygments
    keep_md: no
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
      toc_depth: 3
  keep_md: true  
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      fig.path = "../figures/05_Biodiversity/")
```

# Goals 
Run an analysis on the within-sample biodiversity of the samples in our project. 

1. Calculate the Hill Numbers of the samples using the iNEXT package. 
2. Plot and evaluate the interpolated and extrapolated rarefaction curves from iNEXT. 
3. Analyze the Hill numbers as it relates to stations and salinity (in PSU).
4. Test and then answer the scientific question at hand! (Remember, we're doing biology!)

## Inputs 
1. We will need the `phytree_preprocessed_physeq`, which includes a rooted tree (ideally within the archaea!) that we created in `analysis/04B_Phylogenetic_Tree_Inspection.Rmd`. 

## Outputs 

*This is an exciting analysis!!! We finally get to answer one of our scientific questions!*

1. Calculated biodiversity measures (Hill numbers) across every sample. 
2. Figures to include in our scientific paper that visualize the data as it relates to the scientific question.
3. Statistical tests conveying the measured and quantified changes and patterns in biodiversity.


# Set up the Environment 

## Timing of this script
```{r rmd-start}
# What time did we start running this script? 
start_time <- Sys.time()
```

## Set the seed 
```{r set-seed}
set.seed(20020206)
```

## Load Packages & Colors 
```{r load-packages-colors}
# install packages for stats
#install.packages("rstatix")
#install.packages("ggpubr")

pacman::p_load(tidyverse, devtools, patchwork, iNEXT, phyloseq,
               # packages for stats
               ggpubr, rstatix, install = FALSE)

# Load in colors for plotting 
source("code/colors.R")
```


# Load Data
```{r load-data}
# Load your midpoint-rooted phyloseq object
load("data/04_PhylogeneticTree/midpoint_rooted_physeq.RData")

# Rename for consistency with rest of the script
physeq <- midpoint_rooted_physeq

# Take a quick look
physeq

# Check minimum sequencing depth
min(sample_sums(physeq))

# Extract sample metadata
metadata_df <- 
  sample_data(physeq) %>%
  data.frame()

# View metadata
glimpse(metadata_df)

```


# Run iNEXT to get Hill Number Equation
```{r iNEXT-biodiversity}
# Prepare input data for iNEXT: ASVs in rows, samples in columns
iNEXT_input_df <- 
  otu_table(physeq) %>%
  as.matrix()

# If samples are in rows, transpose the matrix
if (!taxa_are_rows(physeq)) {
  iNEXT_input_df <- t(iNEXT_input_df)
  }


iNEXT_input_df <- data.frame(iNEXT_input_df)

# Quick inspection
dim(iNEXT_input_df)
iNEXT_input_df[1:5, 1:5]

sum(rowSums(iNEXT_input_df) == 0)

# Run iNEXT (can take a while!)
#iNEXT_data <- iNEXT(iNEXT_input_df, 
#                    q = c(0, 1, 2),
#                    datatype = "abundance")

# Save iNEXT results to an .RData file
#save(iNEXT_data, file = "data/05_Biodiversity/iNEXT_data.RData")

```

Since iNEXT takes a while to run, we will load in the data that was caluculated before! 

```{r load-iNEXT-data}
load("data/05_Biodiversity/iNEXT_data.RData")

# Inspect 
str(iNEXT_data)
typeof(iNEXT_data)
```







