---
title: "Biodiversity across a salinity gradient"
author: "Mar Schmidt"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document: 
    code_folding: show
    theme: spacelab
    highlight: pygments
    keep_md: no
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
      toc_depth: 3
  keep_md: true  
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      fig.path = "../figures/05_Biodiversity/")
```

# Goals 
Run an analysis on the within-sample biodiversity of the samples in our project. 

1. Calculate the Hill Numbers of the samples using the iNEXT package. 
2. Plot and evaluate the interpolated and extrapolated rarefaction curves from iNEXT. 
3. Analyze the Hill numbers as it relates to stations and salinity (in PSU).
4. Test and then answer the scientific question at hand! (Remember, we're doing biology!)

## Inputs 
1. We will need the `phytree_preprocessed_physeq`, which includes a rooted tree (ideally within the archaea!) that we created in `analysis/04B_Phylogenetic_Tree_Inspection.Rmd`. 

## Outputs 

*This is an exciting analysis!!! We finally get to answer one of our scientific questions!*

1. Calculated biodiversity measures (Hill numbers) across every sample. 
2. Figures to include in our scientific paper that visualize the data as it relates to the scientific question.
3. Statistical tests conveying the measured and quantified changes and patterns in biodiversity.


# Scientific Question:
How does microbiota restoration using RBX2660 affect microbial biodiversity in patients with recurrent Clostridioides difficile infection?

Null Hypothesis:
There is no significant change in gut microbial biodiversity (e.g., richness, evenness) before and after RBX2660 treatment in patients with recurrent C. difficile infection.

Alternative Hypothesis:
RBX2660 treatment increases gut microbial biodiversity (e.g., richness, evenness), as reflected by significant changes in diversity indices such as Hill numbers before and after therapy.



# Set up the Environment 

## Timing of this script
```{r rmd-start}
# What time did we start running this script? 
start_time <- Sys.time()
```

## Set the seed 
```{r set-seed}
set.seed(20020206)
```

## Load Packages & Colors 
```{r load-packages-colors}
# install packages for stats
#install.packages("rstatix")
#install.packages("ggpubr")

pacman::p_load(tidyverse, devtools, patchwork, iNEXT, phyloseq,
               # packages for stats
               ggpubr, rstatix, install = FALSE)

# Load in colors for plotting 
source("code/colors.R")
```


# Load Data
```{r load-data}
# Load your midpoint-rooted phyloseq object
load("data/04_PhylogeneticTree/midpoint_rooted_physeq.RData")

# Rename for consistency with rest of the script
physeq <- midpoint_rooted_physeq

# Take a quick look
physeq

# Check minimum sequencing depth
min(sample_sums(physeq))

# Extract sample metadata
metadata_df <- 
  sample_data(physeq) %>%
  data.frame()

# View metadata
glimpse(metadata_df)

```


# Run iNEXT to get Hill Number Equation
```{r iNEXT-biodiversity}
# Prepare input data for iNEXT: ASVs in rows, samples in columns
iNEXT_input_df <- 
  otu_table(physeq) %>%
  as.matrix()

# If samples are in rows, transpose the matrix
if (!taxa_are_rows(physeq)) {
  iNEXT_input_df <- t(iNEXT_input_df)
  }


iNEXT_input_df <- data.frame(iNEXT_input_df)

# Quick inspection
dim(iNEXT_input_df)
iNEXT_input_df[1:5, 1:5]

sum(rowSums(iNEXT_input_df) == 0)

# Run iNEXT (can take a while!)
#iNEXT_data <- iNEXT(iNEXT_input_df, 
#                    q = c(0, 1, 2),
#                    datatype = "abundance")

# Save iNEXT results to an .RData file
#save(iNEXT_data, file = "data/05_Biodiversity/iNEXT_data.RData")

#write.csv(iNEXT_data$iNextEst$size_based,
#          file = "data/05_Biodiversity/iNEXT_size_based.csv",
#          row.names = FALSE)

```

Since iNEXT takes a while to run, we will load in the data that was caluculated before! 

```{r load-iNEXT-data}
load("data/05_Biodiversity/iNEXT_data.RData")

# Inspect 
str(iNEXT_data)
typeof(iNEXT_data)

```


# Add a new variable group: donor, SIx, RIx (x = days after therapy)
## SI: 6,11,12,13,19,21,23,26,29,31,32,35, RI: 1,2,3,4,5,7,8,9,10,14,17,18,20,24,27,28,33
```{r add-group}
# Load required packages
library(dplyr)
library(readr)
source("code/assign_group.R")

# Read the sample metadata table
metadata <- read_csv("data/metadata.csv")  # Adjust path if needed

# Create Run-to-SampleName mapping table
run_map <- metadata %>%
  select(Run, SampleName) %>%
  filter(!is.na(SampleName))

# Match Assemblage to SampleName
iNEXT_size_df <- iNEXT_data$iNextEst$size_based %>%
  left_join(run_map, by = c("Assemblage" = "Run"))

# Add grouping info using your existing assign_group() function
iNEXT_size_df_grouped <- iNEXT_size_df %>%
  mutate(Group = sapply(SampleName, assign_group))

# Insert into a new iNEXT_data_group object
iNEXT_data_group <- iNEXT_data
iNEXT_data_group$iNextEst$size_based <- iNEXT_size_df_grouped

# Save the new grouped object
save(iNEXT_data_group, file = "data/05_Biodiversity/iNEXT_data_group.RData")

# Print preview
head(iNEXT_data_group$iNextEst$size_based)


```


## `ggiNEXT` Rarefaction Curves 
```{r iNEXT-colors}
# Prepare color mapping dataframe
color_df <- 
  iNEXT_input_df %>%
  colnames() %>%
  data.frame()

# Rename column
colnames(color_df)[1] <- "names"

# Extract SampleName & Group
iNEXT_color_df <- 
  iNEXT_data_group$iNextEst$size_based %>%
  dplyr::select(SampleName, Group) %>%
  distinct() %>%
  left_join(color_df, by = c("SampleName" = "names"))

# Set factor order for consistent legend and coloring
group_levels <- c("Donor", "SI0", "SI7", "SI14", "SI30", "SI60", "SI90", "SI180",
                  "RI0", "RI7", "RI14", "RI30", "RI60", "RI90", "RI180")
iNEXT_color_df$Group <- factor(iNEXT_color_df$Group, levels = group_levels)

# Manually define group colors
group_colors <- c(
   # Donor
  "Donor" = "#000000",

  # SI
  "SI0"   = "#ccebc5",  
  "SI7"   = "#b2df8a",  
  "SI14"  = "#a6cee3",  
  "SI30"  = "#80b1d3",  
  "SI60"  = "#4daf4a",  
  "SI90"  = "#377eb8",  
  "SI180" = "#08519c",  

  # RI
  "RI0"   = "#ffeda0",  
  "RI7"   = "#feb24c",  
  "RI14"  = "#fd8d3c",  
  "RI30"  = "#fc4e2a",  
  "RI60"  = "#e31a1c",  
  "RI90"  = "#bd0026",  
  "RI180" = "#800026",  

  # NA
  "NA" = "#999999"
)
# View check
head(iNEXT_color_df)

```


# Check sample
```{r}
library(dplyr)

iNEXT_data_group$iNextEst$size_based %>%
  filter(!is.na(Group)) %>%
  group_by(Group, m, Order.q) %>%
  summarise(n_samples = n(), .groups = "drop") %>%
  arrange(Group, Order.q, m) %>%
  group_by(Group, Order.q) %>%
  summarise(min_n = min(n_samples),
            max_n = max(n_samples),
            median_n = median(n_samples),
            total_m_points = n())
```


```{r plot-iNEXT-rarefaction}

# Prepare plot data
iNEXT_df <- iNEXT_data_group$iNextEst$size_based %>%
  filter(!is.na(SampleName), !is.na(Group)) %>%
  arrange(Group, SampleName, Order.q, m)

# Set factor levels for proper legend ordering
iNEXT_df$Group <- factor(iNEXT_df$Group, levels = group_levels)

# Plot with one line per sample
ggplot(iNEXT_df, aes(x = m, y = qD, 
                     group = interaction(SampleName, Method), 
                     color = Group, linetype = Method)) +
  geom_line(size = 0.4, alpha = 0.9) +
  facet_wrap(~ Order.q, scales = "free_y") +
  scale_color_manual(values = group_colors) +
  scale_linetype_manual(values = c("Rarefaction" = "solid", "Extrapolation" = "dashed")) +
  scale_x_continuous(limits = c(0, 60000)) +
  theme_bw(base_size = 12) +
  theme(
    legend.position = "right",
    legend.title = element_blank()
  ) +
  labs(
    title = "Rarefaction Curves for All Samples (Per Sample Line)",
    x = "Number of Sequences",
    y = "Effective Number of ASVs"
  )


```


## Manual Rarefaction Curves
```{r manual-rerafaction}
# Check the structure of iNEXT grouped data object
str(iNEXT_data_group)

# Create plotting dataframe: Hill number estimates for each sample across sequencing depth
iNEXT_manual_df <- 
  iNEXT_data_group$iNextEst$size_based %>%
  dplyr::rename(names = Assemblage) %>%  # Rename column for merge
  left_join(metadata_df, by = c("names" = "SampleName")) %>%  # Merge metadata by SampleName
  mutate(Group = factor(Group, levels = group_levels))  # Set factor levels for Group

# Check dimensions and first few rows
dim(iNEXT_manual_df)
head(iNEXT_manual_df)

# Set Group factor levels to control legend order
group_levels <- c(
  "Donor", "NA",              
  "SI0", "SI7", "SI30", "SI60", "SI90", "SI180", 
  "RI0", "RI7", "RI14", "RI30", "RI60", "RI90", "RI180" 
)
iNEXT_manual_df$Group <- factor(iNEXT_manual_df$Group, levels = group_levels)

# Reorder group_colors to match the levels above exactly
group_colors_ordered <- group_colors[group_levels]

# Plot with legend broken into 3 rows and color order matching factor levels
iNEXT_manual_df %>%
  filter(Method == "Rarefaction") %>%  # Only keep rarefaction curves
  ggplot(aes(x = m, y = qD, color = Group, group = names)) + 
  geom_line(size = 0.5) +  # 画线 Draw line
  facet_grid(Order.q ~ Group, scales = "free_y") +  # Facet by q and Group
  scale_color_manual(values = group_colors_ordered) +  # Match color order
  theme_bw() + 
  labs(
    y = "Effective Number of ASVs", 
    x = "Number of Sequences"      
  ) + 
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1), 
    legend.position = "bottom",                        
    legend.title = element_blank(),                  
    legend.text = element_text(size = 9),        
    legend.box = "horizontal",                    
    legend.justification = "center"             
  ) +
  guides(color = guide_legend(nrow = 3, byrow = TRUE))  

```

<span style="color: red;">INTERPRETATION #1: The rarefaction curves for most samples are stable at all Hill numbers (q = 0, 1, 2), indicating sufficient sequencing depth. When q = 0, species richness can be confidently compared between samples. When q = 1 and q = 2, the curves stabilize earlier, indicating robust detection of common and dominant taxa.</span>


# Shapiro-Wilk
```{r shapiro-wilk-richness}
# Check if richness (q = 0) is normally distributed across treatment groups

# Filter iNEXT results to get observed richness values (q = 0)
obs_div_df <- 
  iNEXT_manual_df %>%
  dplyr::filter(Method == "Observed", Order.q == 0)

# Quick inspection of the structure
glimpse(obs_div_df)

# Shapiro-Wilk test for all samples together
richness_shapiro <- shapiro.test(obs_div_df$qD)
print(richness_shapiro)

```


# Statistically Test Richness all groups
```{r statistically-test-richness-all-groups}
obs_rich_df <- 
  iNEXT_manual_df %>%
  filter(Method == "Observed", Order.q == 0, !is.na(Group)) %>%
  droplevels() 

# Kruskal-Wallis
kw_test <- obs_rich_df %>%
  kruskal_test(qD ~ Group)

# Dunn
dunn_result <- obs_rich_df %>%
  dunn_test(qD ~ Group, p.adjust.method = "fdr") %>%
  add_xy_position(x = "Group")

# Plot
rich_ggboxplot <- 
  ggboxplot(obs_rich_df, x = "Group", y = "qD", 
            color = "Group", fill = "Group", alpha = 0.3,
            outlier.shape = NA) + 
  geom_jitter(aes(color = Group), width = 0.2, size = 1.5) + 
  scale_color_manual(values = group_colors) + 
  scale_fill_manual(values = group_colors) + 
  labs(y = "# of ASVs (Richness)", title = "Richness Differences Between Groups") + 
  stat_pvalue_manual(dunn_result, hide.ns = TRUE, tip.length = 0.01) +
  theme_bw() + 
  theme(
    legend.position = "none",
    axis.title.x = element_blank(),
    axis.text.x = element_text(angle = 30, hjust = 1, vjust = 1)
  )
print(rich_ggboxplot)

```


# Statistically Test Richness donor RI0/7 SI0/7
```{r statistically-test-richness}
# Extract observed richness values (q = 0)
obs_rich_df <- 
  iNEXT_manual_df %>%
  dplyr::filter(Group %in% c("Donor", "SI0", "SI7", "RI0", "RI7")) %>% 
  dplyr::mutate(Group = factor(Group, levels = c("Donor", "SI0", "SI7", "RI0", "RI7"))) %>%
  dplyr::filter(Method == "Observed", Order.q == 0)

# Test for significant richness differences across groups (non-parametric)
kw_test <- obs_rich_df %>%
  kruskal_test(qD ~ Group)

print(kw_test)

# Pairwise comparisons using Dunn test with FDR adjustment
dunn_result <- obs_rich_df %>%
  dunn_test(qD ~ Group, p.adjust.method = "fdr") %>%
  add_xy_position(x = "Group")

# Richness boxplot by Group
rich_plot <- 
  ggboxplot(obs_rich_df, x = "Group", y = "qD", 
          color = "Group", fill = "Group", alpha = 0.3,
          outlier.shape = NA) +
  geom_jitter(aes(color = Group), width = 0.2, size = 1.5) + 
  scale_color_manual(values = group_colors) + 
  scale_fill_manual(values = group_colors) + 
  labs(y = "# of ASVs (Richness)", title = "Richness Differences Between Groups") + 
  stat_pvalue_manual(dunn_result, hide.ns = TRUE, tip.length = 0.01, step.increase = 0.01, bracket.nudge.y = 1) +
  theme_bw() + 
  theme(
    legend.position = "none",
    axis.title.x = element_blank(),
    axis.text.x = element_text(angle = 30, hjust = 1, vjust = 1) 
  )
print(rich_plot)

```


# Statistically Test Shannon donor RI0/7 SI0/7
```{r statistically-test-Shannon}
# Extract observed richness values (q = 1)
obs_rich_df <- 
  iNEXT_manual_df %>%
  dplyr::filter(Group %in% c("Donor", "SI0", "SI7", "RI0", "RI7")) %>% 
  dplyr::mutate(Group = factor(Group, levels = c("Donor", "SI0", "SI7", "RI0", "RI7"))) %>%
  dplyr::filter(Method == "Observed", Order.q == 1)

# Test for significant richness differences across groups (non-parametric)
kw_test <- obs_rich_df %>%
  kruskal_test(qD ~ Group)

print(kw_test)

# Pairwise comparisons using Dunn test with FDR adjustment
dunn_result <- obs_rich_df %>%
  dunn_test(qD ~ Group, p.adjust.method = "fdr") %>%
  add_xy_position(x = "Group")

# Shannon plot by Group
shannon_plot <- 
  ggboxplot(obs_rich_df, x = "Group", y = "qD", 
          color = "Group", fill = "Group", alpha = 0.3,
          outlier.shape = NA) +
  geom_jitter(aes(color = Group), width = 0.2, size = 1.5) + 
  scale_color_manual(values = group_colors) + 
  scale_fill_manual(values = group_colors) + 
  labs(title = "Shannon Diversity Comparison (q = 1)", y = "# of Common ASVs (Shannon)") + 
  stat_pvalue_manual(dunn_result, hide.ns = TRUE, tip.length = 0.01, step.increase = 0.01, bracket.nudge.y = 1) +
  theme_bw() + 
  theme(
    legend.position = "none",
    axis.title.x = element_blank(),
    axis.text.x = element_text(angle = 30, hjust = 1, vjust = 1) 
  )
print(shannon_plot)

```


# Statistically Test Simpson donor RI0/7 SI0/7
```{r statistically-test-Simpson}
# Extract observed richness values (q = 2)
obs_rich_df <- 
  iNEXT_manual_df %>%
  dplyr::filter(Group %in% c("Donor", "SI0", "SI7", "RI0", "RI7")) %>% 
  dplyr::mutate(Group = factor(Group, levels = c("Donor", "SI0", "SI7", "RI0", "RI7"))) %>%
  dplyr::filter(Method == "Observed", Order.q == 2)

# Test for significant richness differences across groups (non-parametric)
kw_test <- obs_rich_df %>%
  kruskal_test(qD ~ Group)

print(kw_test)

# Pairwise comparisons using Dunn test with FDR adjustment
dunn_result <- obs_rich_df %>%
  dunn_test(qD ~ Group, p.adjust.method = "fdr") %>%
  add_xy_position(x = "Group")

# simpson plot by Group
simpson_plot <- 
  ggboxplot(obs_rich_df, x = "Group", y = "qD", 
          color = "Group", fill = "Group", alpha = 0.3,
          outlier.shape = NA) +
  geom_jitter(aes(color = Group), width = 0.2, size = 1.5) + 
  scale_color_manual(values = group_colors) + 
  scale_fill_manual(values = group_colors) + 
  labs(y = "# of Dominant ASVs (Simpson)", title = "Simpson Diversity Comparison (q = 2)") + 
  stat_pvalue_manual(dunn_result, hide.ns = TRUE, tip.length = 0.01, step.increase = 0.01, bracket.nudge.y = 1) +
  theme_bw() + 
  theme(
    legend.position = "none",
    axis.title.x = element_blank(),
    axis.text.x = element_text(angle = 30, hjust = 1, vjust = 1) 
  )
print(simpson_plot)

```


## Plot 3 box plots
```{r 3box-plots}
# Combine richness, Shannon, and Simpson diversity plots in one row
rich_plot + shannon_plot + simpson_plot +
  patchwork::plot_annotation(
    title = "Alpha Diversity Comparison Across Five Groups (Donor, SI/RI 0&7)",
    theme = theme(plot.title = element_text(hjust = 0.5, size = 14, face = "bold")))

```

<span style="color: red;">INTERPRETATION #2:  
Richness (q = 0): There were significant differences between groups, indicating a shift in the number of ASVs observed after RBX2660 treatment. Donor samples generally showed higher richness, indicating a more diverse community structure before treatment.

Shannon Diversity (q = 1): Shannon diversity in treated groups (e.g., RI7) was significantly higher than baseline, indicating an increase in microbial complexity and balance.

Simpson Diversity (q = 2): The higher Simpson diversity in donor and treated groups indicates that these communities are not dominated by a few taxa, reflecting a more even distribution.</span>


<span style="color: red;">INTERPRETATION #3: 
To statistically test the observations in Interpretation #2:

We applied Kruskal-Wallis tests for each Hill number (q = 0, 1, 2) to assess group-level differences;

When global significance was detected, we performed Dunn post hoc tests to identify specific group comparisons;

All p-values were adjusted for multiple comparisons using the FDR (False Discovery Rate) method;

We visualized the results with boxplots annotated with significance levels.

We chose non-parametric tests (Kruskal-Wallis and Dunn) because normality tests (e.g., Shapiro-Wilk) showed the diversity data were not normally distributed, making ANOVA inappropriate.</span>


# Final info for Reproducibility 

## Check Render Time
```{r stop-time}
# Take the time now that we are at the end of the script
end_time <- Sys.time()
end_time 

# Echo the elapsed time
elapsed_time <- round((end_time - start_time), 3)
elapsed_time
```

## Session Information

```{r session-info}
# Ensure reproducibility with package version information
devtools::session_info()
```














